---
title: "Divvy and Data Visualisation"
author: "Lee Zong Han"
output: html_document
date: "2023-01-28"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### ***Disclaimer***

*This analysis is based on the Divvy case study "'Sophisticated, Clear, and Polished': Divvy and Data Visualization" written by Kevin Hartman (found here: [https://artscience.blog/home/divvy-dataviz-case-study).](https://artscience.blog/home/divvy-dataviz-case-study).)*

*The project below was done as part of the Google Data Analytics Certificate.*

## Purpose

This R Markdown is a journal for recording my thought processes and learning. Please feel free to follow along with my journey in tackling this case study.

## Problem Statement

The task is to enable the marketing analysis team to better understand "**In what ways do members and casual riders use Divvy bikes differently?**"

Historical data from Q2 2019 to Q1 2020 was provided for analysis.

## Getting into the actual work process

### Installation of Packages and Environment Preparation

For data wrangling and analysis.

```{r}
library(tidyverse)  #helps wrangle data
library(lubridate)  #helps wrangle date attributes
library(ggplot2)  #helps visualize data
```

For setting up directories.

```{r}
getwd() #displays my working directory
setwd("C:/Users/Lee Zong Han/Desktop/Google Data Analytics Cert/Case Study 1/Divvy")
```

For loading data into environment.

```{r}
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1.csv")
```

### Data Wrangling

Before merging the data sets from the 4 quarters together, I have to check if the columns are synchronised so that the rows can be properly stacked.

Using script below, I saw that the columns in the 4 quarters did not match. I followed the naming convention set in Q1 2020, assuming that this will be the table format moving forward for Divvy.

```{r}
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
```

For renaming the columns in Q2 2019, Q3 2019 and Q4 2019 respectively. Check the tibbles if the columns were correctly renamed.

```{r}
(q2_2019 <- rename(q2_2019
                   ,ride_id = "01 - Rental Details Rental ID"
                   ,rideable_type = "01 - Rental Details Bike ID" 
                   ,started_at = "01 - Rental Details Local Start Time"  
                   ,ended_at = "01 - Rental Details Local End Time"  
                   ,start_station_name = "03 - Rental Start Station Name" 
                   ,start_station_id = "03 - Rental Start Station ID"
                   ,end_station_name = "02 - Rental End Station Name" 
                   ,end_station_id = "02 - Rental End Station ID"
                   ,member_casual = "User Type"))

(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))

(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid 
                   ,started_at = start_time  
                   ,ended_at = end_time  
                   ,start_station_name = from_station_name 
                   ,start_station_id = from_station_id 
                   ,end_station_name = to_station_name 
                   ,end_station_id = to_station_id 
                   ,member_casual = usertype))
```

Check for inconsistency in the data types. Incompatible data will not stack correctly.

```{r}
str(q2_2019)
str(q3_2019)
str(q4_2019)
str(q1_2020)
```

Upon checking, ride_id and rideable_type were found to be in a non-character format. Convert data types to character for proper stacking.

```{r}
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
q4_2019 <-  mutate(q4_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)) 
```

Combine all quarter data sets into a single data frame to facilitate analysis.

```{r}
all_trips <- bind_rows(q2_2019, q3_2019, q4_2019, q1_2020)
```

Remove all latitude, longitude, gender and birth year fields as these were removed in Q1 2020.

```{r}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender, "01 - Rental Details Duration In Seconds Uncapped", "05 - Member Details Member Birthday Year", "Member Gender", "tripduration"))
```

### Cleaning Data to Prepare for Analysis

First, the new data frame that has been created was inspected.

```{r}
colnames(all_trips)  #List of column names
nrow(all_trips)  #How many rows are in data frame?
dim(all_trips)  #Dimensions of the data frame?
head(all_trips)  #See the first 6 rows of data frame.  Also tail(all_trips)
str(all_trips)  #See list of columns and data types (numeric, character, etc)
summary(all_trips)  #Statistical summary of data. Mainly for numerics
```

Below are some findings and how we are going to handle them:

#### Problem 1 - Standardising Labels

The "member_casual" column contains multiple labels for members ("member" and "Subscriber") and casual riders ("Customer" and "casual"). To simplify the analysis, these labels will be consolidated into two distinct categories.

We will use the naming convention from the Q1 2020 data. In the "member_casual" column, replace "Subscriber" with "member" and "Customer" with "casual".

```{r}
table(all_trips$member_casual) #Check how many observations are present for each usertype
all_trips <-  all_trips %>% #Reassign the labels
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))
table(all_trips$member_casual) #Check to ensure usertypes were properly reassigned
```

#### Problem 2 - Creating Categories for Aggregation

The data currently only allows for analysis at a ride-type level, which is too detailed. To increase the possibilities of data aggregation, additional columns such as day, month and year will be added to the data.

```{r}
all_trips$date <- as.Date(all_trips$started_at) #Default format is yyyy-mm-dd
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

#### Problem 3 - Creating Duration Data

There is no "tripduration" column in the Q1 2020 data. Adding a "ride_length" column to the data frame can assist with further analysis on bike usage patterns.

```{r}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)
str(all_trips)
is.factor(all_trips$ride_length) #Check data type
all_trips$ride_length <- as.numeric(as.character(all_trips$ride_length)) #Convert "ride_length" from factor to numeric so we can run calculations on the data
is.numeric(all_trips$ride_length) #Check data type
```

#### Problem 4 - Removing Bad Data

Some rides have negative values in the "tripduration" column and some rides were taken out for circulation for Quality Control. To avoid skewing the data, these rides will be removed.

A new data frame all_trips_v2 will be created and worked from moving forward.

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]
```

## Conducting Descriptive Analysis

### Understanding the Data

#### Basic Statistics

A simple analysis of the data can be achieved by looking at the mean, median, maximum and minimum ride length.

```{r}
mean(all_trips_v2$ride_length) #straight average (total ride length / rides)
median(all_trips_v2$ride_length) #midpoint number in the ascending array of ride lengths
max(all_trips_v2$ride_length) #longest ride
min(all_trips_v2$ride_length) #shortest ride
```

The below script can achieve the same results as the code above with one line.

```{r}
summary(all_trips_v2$ride_length)
```

#### Making Comparisons Within the Data

By creating new categories for aggregation earlier, we can partition the data by these categories to better understand inter-group trends.

For example, we can understand the ride length **difference between members and casual riders** by partitioning by "member_casual".

```{r}
# Compare members and casual users
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual, FUN = min)
```

#### Ordering Unique Values Within a Categorical Value

To understand the average ride time on each day for each user type, we can partition the average ride length by:

1.  Day of Week
2.  Type of Rider

```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
```

However, we notice that the days are not displayed in order. We can clean this up by making use of levels.

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday")) #Order days

aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean) #Check order
```

We can also analyse ridership by weekdays only. This is done by creating a new field.

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>%  #Create weekday field
  group_by(member_casual, weekday) %>%  #Groups by user type and weekday
  summarise(number_of_rides = n() #Calculates the number of rides
            ,average_duration = mean(ride_length)) %>% #Calculates the average duration
  arrange(member_casual, weekday)	#Sorts data
```

## Interpreting the Data

### Data Visualisation

Visualisations can make it easier to compare and contrast different data points and to understand the relationships between variables.

### Linking Back to the Problem Statement

With the data, we can draw insights about the behaviour of bike users based on observations:

#### Metric 1 - Frequency of Rides

```{r}
all_trips_v2 %>% #Visualise number of rides by user type
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
    summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = number_of_rides, fill = member_casual)) +
  geom_col(position = "dodge")
```

[Observation]{.underline}

From the plot, we can see:

1.  The frequency of rides by member riders on weekdays are greater than on weekends.
2.  The frequency of rides by casual riders on weekdays are fewer than on weekends.

[Insight]{.underline}

With this trend, we can suspect that:

1.  Member ridership is correlated to the need for riding on weekdays. Needs for riding on weekdays might include:
    -   Scheduled activities like **work, school or errands** etc.
2.  Casual ridership is correlated to the need for riding on weekends. The demand for this group of riders could stem from **leisure riding** on the weekends.

[Further Study]{.underline}

More data could be collected on the behaviours of riders via surveys to support the insights derived above.

#### Metric 2 - Average Duration of Rides

```{r}
all_trips_v2 %>% 
  mutate(weekday = wday(started_at, label = TRUE)) %>% 
  group_by(member_casual, weekday) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, weekday)  %>% 
  ggplot(aes(x = weekday, y = average_duration, fill = member_casual)) +
  geom_col(position = "dodge")
```

[Observation]{.underline}

From the plot, we can see:

1.  The average duration of each ride is consistently longer by casual riders compared to member riders for every day of the week.

[Insight]{.underline}

From the limited data, it is difficult to identify why the length of rides for casual and member riders differ so much. More information such as "Reason for riding" could be collected to further study this observation.

## Sharing the Work

After our initial analysis, we can export this data or this RMarkdown document for the marketing team to study.

```{r}
counts <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$member_casual + all_trips_v2$day_of_week, FUN = mean)
write.csv(counts, file = "C:/Users/Lee Zong Han/Desktop/Google Data Analytics Cert/Case Study 1/Divvy/avg_ride_length.csv")
```

Another way of sharing insights is with the use of business intelligence tools such as Tableau.

## Where to Find My Projects?

This analysis will be hosted on both my Github and Tableau Public accounts.

Github: <https://github.com/zonghanleezh/PortfolioProjects>

Tableau Public: <https://public.tableau.com/app/profile/lee.zong.han>
